---
params:
  myTitle: 'Water Quality in New Zealand'
  mySubTitle: 'Exploring "official" data'
title: "`r params$myTitle`"
subtitle: "`r params$mySubTitle`"
author: "Ben Anderson (b.anderson@soton.ac.uk `@dataknut`)"
date: 'Last run at: `r Sys.time()`'
output:
  bookdown::html_document2:
    fig_caption: yes
    number_sections: yes
    toc: yes
    collapsed: FALSE
    toc_float: yes
    highlight: tango
    code_folding: hide
bibliography: '`r path.expand("~/bibliography.bib")`' 
---

```{r knitrSetUp, include=FALSE}

# Set knitr options
knitr::opts_chunk$set(echo = FALSE,          # echo code so reader can see what is happening
                      warning = FALSE,
                      message = FALSE,
                      fig_caption = TRUE,
                      fig_height = 6,        # default, make it bigger to stretch vertical axis
                      fig_width = 8,         # full width
                      tidy = TRUE)           # tidy up code in case echo = TRUE

```


```{r codeSetup, include=FALSE}

# Load Packages ----
library(dkUtils)

myLibs <- c("data.table",
            "forcats",
            "ggplot2",
            "here",
            "kableExtra",
            "lubridate",
            "plotly",
            "skimr", 
            "viridis")

dkUtils::loadLibraries(myLibs)              # Load script specific packages

# Project Settings ----
projLoc <- here::here()

dPath <- "~/Data/NZ_mfe/waterQuality/"
dataFile <- "mfe-river-water-quality-trends-20082017-19982017-and-19902017-CSV/river-water-quality-trends-20082017-19982017-and-1990-2017-raw-data.csv"

wqURL <- "https://data.mfe.govt.nz/data/category/fresh-water/"
  

# Adjust knitr options if required
knitr::opts_chunk$set(echo = TRUE)

# Log compile time:
startTime <- proc.time()

# Functions ----
makeTilePlot <- function(dt,yvar,byvar){
  p <- ggplot2::ggplot(dt, aes(x = ba_date, 
                               y = get(byvar), 
                               fill = get(yvar)
                               )
                       ) +
    geom_tile() +
    scale_fill_continuous(low = "green", high = "red") +
    labs(x = "Date")
  return(p)
}

makeLinePlot <- function(dt,yvar,byvar){
  p <- ggplot2::ggplot(dt, aes(x = ba_date, 
                               y = get(yvar),
                               colour = get(byvar)
                               )
                       ) +
    geom_line() +
    labs(x = "Date")
  return(p)
}
```

# About

## Contributions

Please note that authorship is alphabetical. Contributions are listed below - see [github](https://github.com/CfSOtago/waterQual/commits/master) for details and who to blame for what :-).

Ben Anderson (b.anderson@soton.ac.uk `@dataknut`)

## Code

 * Repo: https://github.com/CfSOtago/waterQual/

## Citation

If you wish to refer to any of the material from this report please cite as:

 * Anderson, B., (`r format(Sys.time(), format = "%Y")`) `r params$myTitle`: `r params$mySubTitle` , University of Southampton: Southampton, UK.

Report circulation:

 * Public.
 
Report purpose: 
 
 * to explore `official` New Zealand Water Quality data

This work has received funding from the European Union’s Horizon 2020 research and innovation programme under the [Marie Skłodowska-Curie](http://ec.europa.eu/research/mariecurieactions/index_en.htm) grant agreement No 700386 ([SPATIALEC](http://www.energy.soton.ac.uk/tag/spatialec/)).

This work is (c) `r format(Sys.time(), format = "%Y")` the University of Southampton.
 
# Introduction

# MfE data

Data source: `r wqURL`

Data file: `r dataFile`

```{r loadTrend}

df <- paste0(dPath, dataFile)

mfeWqDT <- data.table::fread(df)
```

Overall there are:

 * `r uniqueN(mfeWqDT$sID)` sites

This isn't really trend or raw data - it is a single record per site with indicator of change via a trend slope coefficient or a single % change value. We were looking for the data used to create the time-series plots in (for example) https://www.lawa.org.nz/explore-data/otago-region/river-quality/manuherikia-river/manuherikia-at-ophir/

There are also data quality & missing data issues (see Statistical Annex). It's noticeable for example that, irrespective of the indicator, over half of stations were thought to have insufficient data to determine trend direction. This was true of 70% of stations for [ECOLI](https://www.stats.govt.nz/indicators/river-water-quality-escherichia-coli) and 78% for the Macroinvertebrate Community Index ([MCI](https://www.stats.govt.nz/indicators/river-water-quality-macroinvertebrate-community-index)).

```{r summariseQuality}
tq <- table(mfeWqDT$TrendCategory, mfeWqDT$npID)

kableExtra::kable(addmargins(tq), caption = "Trend category counts") %>%
  kable_styling()

kableExtra::kable(100*prop.table(tq,2), digits = 2,
                  caption = "Trend category %") %>%
  kable_styling()
```


```{r aggTrends, fig.height = 8, fig.cap="Test data values by indicator"}
plotDT <- mfeWqDT[, .(meanAnnualSenSlope = mean(AnnualSenSlope, na.rm = TRUE)),
               keyby = .(npID, TrendCategory)]
ggplot2::ggplot(plotDT, aes(x=TrendCategory, y = meanAnnualSenSlope, 
                            group = npID, fill = npID)) + 
  geom_col(position = "dodge")

plotDT <- mfeWqDT[, .(meanPercent_annual_change = mean(Percent_annual_change, na.rm = TRUE)),
               keyby = .(npID, TrendCategory)]
ggplot2::ggplot(plotDT, aes(x=TrendCategory, y = meanPercent_annual_change, 
                            group = npID, fill = npID)) + 
  geom_col(position = "dodge")

```

Now try to plot individual trends. This is nasty...

Ideally we want to sort by value within pollutant and 'fade out' the ones with insufficient data. Colour coding needs to be careful - for some such as MCI an increase is 'good'.

```{r plotSites, fig.height = 15, fig.cap="Site level trend plots"}
p <- ggplot2::ggplot(mfeWqDT, aes(y = sID,
                               x = Percent_annual_change, 
                               alpha = forcats::fct_rev(TrendCategory), # reverse so lightest alpha == least data
                               colour = forcats::fct_rev(TrendCategory)
                               )
                     ) +
  geom_vline(xintercept = 0, colour = "red") +
  geom_point() +
  guides(alpha = guide_legend(title = "Trend and quality code")
         ) +
  guides(colour = guide_legend(title = "Trend and quality code")
         ) +
  scale_color_viridis(discrete = TRUE) +
  theme(legend.position = "bottom") +
  facet_grid(. ~ npID)

p
```

# LAWA data

So this is the trend data we were after...

 * https://www.lawa.org.nz/download-data/
 
Also downloadable from:

 * https://statisticsnz.shinyapps.io/river_water_quality_phosphorus/ (for example)
 
```{r setLawa}

f <- "LAWA Rivers_DownloadWQRawData_Oct2018.csv"
p <- "~/Data/NZ_LAWA/waterQuality/"

df <- paste0(p, f)

LAWA_DT <- data.table::fread(df)
```

Using:

 * `r f`

```{r checkLawa, fig.height = 10}

names(LAWA_DT)
LAWA_DT$Ownership <- NULL # not used
LAWA_DT$License <- NULL # not used
LAWA_DT$Disclaimer <- NULL # not used

LAWA_DT[, RawValue := as.numeric(RawValue)]
LAWA_DT[, region.site := paste0(Region, ".", Site)]

LAWA_DT[, ba_date := lubridate::dmy(Date)]

h <- head(LAWA_DT)

kableExtra::kable(h, caption = "First 6 rows of LAWA data") %>%
  kable_styling()

p <- makeTilePlot(LAWA_DT[parameter == "ECOLI"], yvar = "Value", byvar = "Site")
p + labs(y = "ECOLI", caption = "Observations for E.Coli") + 
  theme(legend.position = "bottom") +
  guides(fill=guide_legend(title="E.Coli"))

p <- makeTilePlot(LAWA_DT[parameter == "NH4"], yvar = "Value", byvar = "Site")
p + labs(y = "NH4", caption = "Observations for NH4") + 
  theme(legend.position = "bottom") +
  guides(fill=guide_legend(title="NH4"))
```


```{r lawaSummary}

st <- summary(LAWA_DT)

kableExtra::kable(st, caption = "Summary of LAWA data") %>%
  kable_styling()

```

```{r testDataPlotly, fig.height = 10, fig.cap="Test E.Coli data values by date and site"}
# looks like daily data with gaps
p <- makeLinePlot(LAWA_DT[parameter == "ECOLI"],
                  yvar = "Value",
                  byvar = "Site" )

p <- p + labs(y = "E. coli", caption = "E. coli") +
  guides(colour=guide_legend(title="E. coli")) +
  theme(legend.position = "none") +
  facet_grid(Region ~ .)

p

```

# Statistical Annex

## MfE 'raw' data

```{r skimMfE}
skimr::skim(mfeWqDT)
```

```{r trendNotes}

tn <- mfemfeWqDT[!is.na(AnalysisNote) & AnalysisNote != "ok", .(sID, npID, AnalysisNote)]

kableExtra::kable(tn, caption = "Analysis notes")

```


## LAWA 'raw' data

```{r skimLAWA}
skimr::skim(LAWA_DT)
```


# Runtime

Report generated using [knitr](https://cran.r-project.org/package=knitr) in [RStudio](http://www.rstudio.com) with `r R.version.string` running on `r R.version$platform` (`r Sys.info()[3]`).

```{r check runtime}
t <- proc.time() - startTime

elapsed <- t[[3]]
```

Analysis completed in `r elapsed` seconds ( `r round(elapsed/60,2)` minutes).

R packages used:

 * data.table - [@data.table]
 * forcats - [@forcats]
 * ggplot2 - [@ggplot2]
 * here - [@here]
 * kableExtra - [@kableExtra]
 * lubridate - [@lubridate]
 * skimr - [@skimr]
 * viridis - [@viridis]
            
# References

